\Title

### 2013-02-12 synmut.tex
edits throughout, reorganized discussion and supplement

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{fix}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which need to avoid recognition by
neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are strongly selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes since these processes often
depend on RNA folding.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.

Even in absense of important RNA structure synonymous codons don't
evolve completely neutrally but that some codons are favored over others in many
species \citep{plotkin_synonymous_2011}.
Recent studies have shown that genetically engineered HIV strains with
altered codon usage can in some cases produce more viral protein, but in general
replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}. Positive
selection through the host's immune system, however, is restricted to
changes in the amino acid sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (see methods). The data set from
\citet{shankarappa_consistent_1999,liu_selection_2006} is restricted to the
C2-V5 region of \env, while \citet{bunnik_autologous_2008} covers the
majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t_i$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t_f$. Plotting these fixed, lost, and polymorphic fraction against
the time interval $t_f-t_i$, we see that most synonymous mutations segregate for
roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{Time series of frequencies of synonymous (A) and non-synonymous (B)
derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) Time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is below the neutral
expectation indicated by the diagonal line. This suppression is not
oberserved in other parts of {\it env} or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible {\it a priori} explanation for lack of fixation of synonymous
mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
which is deleterious to the virus.
It has been suggested early on that parts of the viral genome that have the
potential to form RNA stems are better conserved than the
remainder~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011}.

Recently, the propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles are lost (blue
) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
Note that, because of small differences in base pairing between the reference
strain NL4-3 and each patient's initial consensus sequence, this analysis is not
expected to be very sensitive, but very conservative.

To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved streches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations have
high SHAPE reactivities.
Panel A) Synonymous substitions have
substantially higher SHAPE reactivities than mutations that reach frequencies above 15\%, but are
susequently lost (restricted to the regions V1-V5$\pm 100bp$ from each). This difference
is quantified by a cumulative histogram ($p=0.002$, Kolmogorov-Smirnov two 
sample test). Mutations that are never
observered show an intermediate distribution of SHAPE values.
Panel B) Fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions, that
harbor conserved RNA stems. As expected, fixation probability is lower inside
the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency -- at least in some regions of the genome. The region of \env~in
which we observe deleterious mutations at high frequency undergoes frequent
adaptive changes to evade recognition by neutralizing 
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the 
limited amount of recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
Hitchhiking is  apparent in \FIG{aft}, that shows that many mutations
change rapidly in frequency as a flock.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from uninformed prior distributions in the indicated window (see methods).}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010, Asquith}.
\comment{Asquith??}

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations. 
We quantify synonymous diversity,$P_\text{interm}$, as the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.
To quantify the depression of the fixation probability, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). If no fixation happens,
the area will be $-0.5$; if every mutation fixes, we will measure 
$A= 0.5$. In HIV, we find $A_\text{syn} \sim -0.2$  for
synonymous changes and $A_\text{nonsyn} \sim 0$ for nonsynonymous changes.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ARTICLE ABOUT FATE OF SYNONYMOUS MUTATIONS IN HIV
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[rmp, twocolumn]{revtex4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath,amsfonts,amssymb,eucal,eurosym,textcomp}
\usepackage{color}
\usepackage{graphicx}
\usepackage[caption=false]{subfig}
\usepackage{natbib}
\usepackage{pslatex}
\usepackage[colorlinks,linkcolor=red,citecolor=red]{hyperref}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./figures/}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newcommand{\comment}[1]{\textit{\textcolor{red}{#1}}}
\newcommand{\mut}{\mu}
\newcommand{\mfit}{\langle F\rangle}
\newcommand{\mexpfit}{\langle e^{F}\rangle}
\newcommand{\pfix}{P_{\mathrm{fix}}}
\newcommand{\ox}{r}
\newcommand{\co}{\rho}
\newcommand{\gt}{g}
\newcommand{\locus}{s}
\newcommand{\locuspm}{t}
\newcommand{\OO}{\mathcal{O}}
\newcommand{\rev}{\textit{rev}}
\newcommand{\FIG}[1]{Fig.~\ref{fig:#1}}
\newcommand{\FIGS}[2]{Figs.~\ref{fig:#1} and~\ref{fig:#2}}
\newcommand{\env}{\textit{env}}
\newcommand{\shankaregion}{C2-V5}
\newcommand{\timedependence}{4}
\newcommand{\withinepi}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\renewcommand{\thesubfigure}{\Alph{subfigure}}
\newcommand{\Author}{Fabio~Zanini and Richard~A.~Neher}
\newcommand{\Title}{Deleterious synonymous mutations hitchhike to high frequency in HIV \env~evolution}
\newcommand{\Keywords}{{HIV}, {synonymous}, {population genetics}}
\hypersetup{pdfauthor={\Author}, pdftitle={\Title}, pdfkeywords={\Keywords}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\title{\Title}
\author{\Author}
\date{\today}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

HIV evolves rapidly within a single host during the course of the infection.
This evolution is driven by strong selection imposed by the host immune system
via killer T cells (CTLs) and neutralizing antibodies
(ABs)~\citep{rambaut_causes_2004} and facilitated by the high
mutation rate of HIV~\citep{mansky_lower_1995,abram_nature_2010}. When the host
develops a CTL or AB response against a particular HIV epitope, mutations in the viral genome that
reduce or prevent recognition of the epitope frequently emerge. Escape mutations
in epitopes targeted by CTLs typically evolve during early infection and spread
rapidly through the population~\citep{mcmichael_immune_2009}. During chronic
infection, the most rapidly evolving part of the HIV genome are the so called
variable loops of the envelope protein gp120, which change to avoid recognition
by neutralizing ABs.  Mutations in \env, the gene encoding gp120, spread
through the population within a few months (see \figurename~\ref{fig:aftnonsyn}).
Consistent with this time scale, it is found that serum from a
particular time typically neutralizes virus extracted more than 3-6 month
earlier \citep{richman_rapid_2003}.

These escape mutations are selected for their effect on the amino acid
sequence of the viral proteins. Conversely, synonymous mutations are commonly
used as approximately neutral markers in studies of viral evolution. Neutral
markers are very useful since their dynamics can be compared to
that of putatively functional sites to detect purifying or directional selection
\citep{Bhatt:2011p43255,Hurst:2002p32608,Chen:2004p22606}.
In addition to maintaining protein function and avoiding the adaptive immune
recognition, however, the HIV genome has to ensure efficient processing and translation,
nuclear export, and packaging into the viral capsid: all these processes operate at the RNA
level and are sensitive to synonymous changes since these processes often
depend on RNA folding.
For example, the HIV \rev{} response element (RRE) in \env{} enhances nuclear export of
full length or partially spliced viral transcripts~\citep{fernandes_hiv-1_2012}.
Another well studied case is the interaction between viral reverse transcriptase, viral ssRNA, and the host
tRNA$^\text{Lys3}$: the latter is required for priming reverse transcription
(RT) and is bound by a pseudoknotted RNA structure in the viral 5'
untranslated region~\citep{barat_interaction_1991, paillart_vitro_2002}.

Even in absense of important RNA structure synonymous codons don't
evolve completely neutrally but that some codons are favored over others in many
species \citep{plotkin_synonymous_2011}. 
Recent studies have shown that genetically engineered HIV strains with altered codon usage can in some cases produce more viral protein, but in general replicate less efficiently~\citep{ngumbela_quantitative_2008,
li_codon-usage-based_2012,keating_rich_2009}.
Codon-deoptimization has been suggested as attenuation strategy for polio and 
influenza~\citep{mueller_live_2010,coleman_virus_2008}. Purifying
selection beyond the protein sequence is therefore expected
\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011} and it has been
shown that rates of evolution at synonymous sites
vary along the HIV genome \citep{mayrose_towards_2007}.
Positive selection through the host's immune system, however, is restricted to
changes in the amino acid sequence.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous polymorphisms in \env, C2-V5, are mostly deleterious}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We study the dynamics and fate of synonymous mutations more quantitatively by
analyzing data from 7 patients from
\citet{shankarappa_consistent_1999,liu_selection_2006} and 3 patients from
\citet{bunnik_autologous_2008} (patients whose viral population was
structured were excluded from the analysis; see methods).
The former data set from is restricted to the
C2-V5 region of \env, while the data from \citet{bunnik_autologous_2008} covers
the majority of \env. Considering all mutations in a
frequency interval around $\nu_0$ at some time $t$, we calculate the fraction
that is found at frequency 1, at frequency 0, or at intermediate frequency at
later times $t+\Delta t$. Plotting these fixed, lost, and polymorphic fraction
against the time interval $\Delta t$, we see that most synonymous mutations
segregate for roughly one year and are lost much more frequently than expected (panel
\ref{fig:fixp1}). The long-time probability of fixation versus extinction of
synonymous mutations is shown as a function of the initial frequency $\nu_0$ in
panel~\ref{fig:fixp2} (red line). Restricted to the region C2-V5, we find that
$\pfix$ of synonymous mutations is far below the neutral expectation.
Outside of C2-V5, using data from \citet{bunnik_autologous_2008} only, no such
reduction in $\pfix$ is found. Restricted to the C2-V5 area, the data from
\citet{bunnik_autologous_2008} is fully compatible with data from
\citet{shankarappa_consistent_1999}. The nonsynonymous mutations seem to follow
more or less the neutral expectation (blue line) -- a point to which we will come back below.

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_syn_p1.pdf}
\label{fig:aftsyn}}\\
 \subfloat{\includegraphics[width=\linewidth]{Shankarappa_allele_freqs_trajectories_nonsyn_p1.pdf}
\label{fig:aftnonsyn}}
\caption{\comment{I would like p2 or p10 better here} Time series of frequencies
of synonymous (A) and non-synonymous (B) derived alleles in \env, C2-V5, from patient
1~\cite{shankarappa_consistent_1999}.
While many nonsynonymous mutations  fix, few synonymous
mutations do even though they are frequently observed at intermediate
frequencies. Colors indicate the position of the site along the C2-V5 region
(blue to red). Inset: the fixation probability $P_\text{fix}(\nu)$ of a neutral mutation
is simply the likelihood that the future common ancestor is currently carrying
it, i.e., its frequency $\nu$.}
\label{fig:aft}
\end{center}
\end{figure}

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_times}
\label{fig:fixp1}}\\
\subfloat{\includegraphics[width=0.9\linewidth]{fixation_probabilities}
\label{fig:fixp2}}
\caption{Fixation and loss of synonymous mutations.
Panel A) shows the time
course of loss and fixation of synonymous mutations observed in a frequency interval $\nu_0$. 
In each frequency interval, the  fraction of synonymous
mutations that ultimately fix is the fixation probability conditional on the
initial frequency.
Panel B) Fixation probability of derived synonymous
alleles as a function of $\nu_0$ in C2-V5 is reduced relative to the neutral
expectation indicated by the diagonal line. This suppression is not
observed in other parts of \env~or for nonsynonymous mutations.
The horizontal error bars on the abscissa are bin sizes, the vertical ones the
standard deviation after 100 patient bootstraps of the data. Data from
Refs.~\cite{shankarappa_consistent_1999,liu_selection_2006, bunnik_autologous_2008}.}
\label{fig:fixp}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Synonymous mutations in C2-V5 tend to disrupt conserved RNA stems}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
One possible explanation for lack of fixation of synonymous
mutations in C2-V5 are secondary structures in the viral RNA, the disruption of
which is deleterious to the virus
~\citep{forsdyke_reciprocal_1995,snoeck_mapping_2011,sanjuan_interplay_2011}.

The propensity of nucleotides of the HIV genome to form base pairs has
been measured using the SHAPE assay (a biochemical reaction preferentially
altering unpaired bases)~\citep{watts_architecture_2009}. The SHAPE assay has
shown that the variable regions V1 to V5 tend to be unpaired, while the
conserved regions between those variable regions form stems. We partition all
synonymous alleles observed at intermediate frequencies above 10-15\% depending
on their final destiny (fixation or extinction). Subsequently, we align our
sequences to the reference NL4-3 strain used in
ref.~\citep{watts_architecture_2009} and assign them SHAPE reactivities. As
shown in \FIG{SHAPEA}, the reactivities of fixed
alleles (red histogram) are systematically larger than of alleles that are lost
(blue ) (Kolmogorov-Smirnov test, $P\approx 0.002$).
In other words, alleles that are likely to break RNA helices are also more
likely to revert and finally be lost from the population. The
average over all mutations that are not observed (green) lies between the
those that fix and those that get lost.
Note that, that this analysis will be sensitive only at position where the base
pairing pattern of NL4-3 agrees with that of each patient's initial
consensus sequence.

To test the hypothesis that mutations in C2-V5 are lost because they break stems in the
conserved stretches between the variable loops, we consider mutations in
variable loops and conserved parts separately. The biggest depression in
fixation probability is observed in the conserved stems, while the variable
loops show little deviation from the neutral signature, see \FIG{SHAPEB}. This
is consistent with important stem structures in conservered regions between loops.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[height=0.46\linewidth]{reactivities_histograms_syn}
\label{fig:SHAPEA}}
\subfloat{\includegraphics[height=0.46\linewidth]{fixation_probabilities_VnonV}\label{fig:SHAPEB}}
\caption{Permissible synonymous mutations tend to be unpaired.
Panel A) shows the distribution of SHAPE reactivities among synonymous mutations
that fix, mutations that reach frequencies above 15\%, but
are susequently lost, and random mutations (all categories are restricted to the
regions V1-V5$\pm 100$bp).
Fixed mutations tend to have higher higher SHAPE reactivities, corresponding to
less base pairing ($p=0.002$, Kolmogorov-Smirnov two sample test on the
corresponding cumulative distribution).
Mutations that are never observered show an intermediate distribution of SHAPE values.
Panel B) shows the fixation probability of synonymous mutations in C2-V5
separately for variable loops and the connecting conserved regions that
harbor conserved RNA stems. As expected, the fixation probability is lower
inside the conserved regions. Data from Refs.~\cite{shankarappa_consistent_1999,
bunnik_autologous_2008, liu_selection_2006}.}
\label{fig:SHAPE}
\end{center}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Deleterious mutations are brought to high frequency by hitchhiking}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
While the observation that some fraction of synonymous mutations is deleterious
is not unexpected, it seems odd that we observe them at high population
frequency and that the fixation probability is reduced only in parts
of the genome.
This regions, however, undergoes frequent
adaptive changes to evade recognition by neutralizing 
antibodies~\cite{williamson_adaptation_2003,richman_rapid_2003}. Due to the 
limited amount of recombination in HIV~\cite{neher_recombination_2010,batorsky_estimate_2011},
deleterious mutations that are linked to adaptive variants can reach high
frequency. This process is known as hitchhiking \citep{smith_hitch-hiking_1974}
or genetic draft \citep{gillespie_genetic_2000,neher_genetic_2011}.
Hitchhiking is  apparent in \FIG{aft}, that shows that many mutations
change rapidly in frequency as a flock.

@@

\begin{figure}
\begin{center}
\subfloat{\includegraphics[width=\linewidth]{simulations_graduallydel.pdf}
\label{fig:simfixpvar}}\\
\subfloat{\includegraphics[width=\linewidth]{simulations_syn.pdf}
\label{fig:simsfig}}
\caption{Distribution of selection coefficients on synonymous sites. Panel A)
The depression in $P_\text{fix}$ depends on the deleterious effect size 
of the synonymous alleles. This parameter also reduces synonymous
diversity, measured by the probability of a derived allele to be found at
intermediate frequencies $P_\text{interm}$ (first inset).
Panel B) To assess the parameter space that affects synonymous fixation and
diversity, we run many simulations with random parameters for deleterious effect
size, fraction of deleterious synonymous sites, average size of escape
mutations (color, blue ($10^{-2.5}$) to red ($10^{-1.5}$)), and rate of
introduction of new epitopes (marker size, from $10^{-3}$ to $10^{-2}$ per
generation). Mostly simulations with small deleterious effects and relatively
high deleterious fractions of synonymous mutations reproduce synonymous the
fixation probability and diversity observed in patients. Parameters are chosen
from prior distributions uniform in log-space indicated by the red box (see
methods).}
\label{fig:simheat}
\end{center}
\end{figure}

To get a better idea of the range of parameters that are compatible with the
observations and our interpretation, we performed computer simulations of
evolving viral populations assuming a mix of positive and purifying selection
and rare recombination.
For this purpose, we use the simulation package FFPopSim, which includes a
module dedicated to intrapatient HIV evolution~\citep{zanini_ffpopsim:_2012}. 
For each simulation run, we specify the deleterious effect of synonymous
mutations, the fraction of synonymous mutations that are deleterious, the
escape rate of adaptive non-synonymous mutations and the frequency of new
escapes. Note that the escape rate is sum of two factors: (i) the beneficial
effect due to the ability to evade the immune system minus (ii) the fitness
cost of the mutation in terms of structure, stability, etc. Net escape rates
in chronic infections have been estimated to be on the order of 
$s_a = 0.01$ per day~\citep{neher_recombination_2010,Asquith:2006p28003}.

\FIG{simfixpvar} shows simulations results for the fixation probability and the
synonymous diversity for different deleterious effects of synonymous mutations. 
We quantify synonymous diversity,$P_\text{interm}$, as the fraction of sites
with an allele at frequency $0.25 < \nu < 0.75$. The synonymous diversity
observed in patient data is indicated in the figure.
To quantify the depression of the fixation probability, we calculate
the area between the measured fixation probability and the diagonal, which is
the neutral expectation (\FIG{simfixpvar}, lower inset). If no fixation happens,
the area will be $-0.5$; if every mutation fixes, we will measure 
$A= 0.5$. In HIV, we find $A_\text{syn} \approx -0.2$  for
synonymous changes and $A_\text{nonsyn} \approx 0$ for nonsynonymous changes.

In the three simulations shown in \FIG{simfixpvar}, the fixation probability of
synonymous alleles decreases from the neutral expectation ($A_\text{syn} \sim 0$) to zero
($A_\text{syn} \sim -0.5$) as their fitness effect
worsens; the synonymous diversity plummets as well, as deleterious mutations
are selected against.
